---
title: "Review of EPI Info"
subtitle: "StatCalc Calculator"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Source link: [Download EPI Info software](https://www.cdc.gov/epiinfo/support/downloads.html)


## 1. Summary

The *StatCalc* offers analytic computation for pre and post-data collection. It comprises eight modules. Three modules serve as sample size calculators: *Population Survey*, *Cohort or Cross-Sectional*, and *Unmatched Case-Control*. These three modules calculate sample size for proportion.

The input parameters of modules for calculating sample sizes:

* The *Population Survey* input parameters: population size, expected frequency, an acceptable margin of error, design effect, and the number of clusters. 

* The *Cohort or Cross-Sectional* input parameters: two-sided confidence level, power, a ratio of unexposed to exposed, percent outcome in the unexposed group, and one of the three remaining inputs. These three inputs are risk ratio, odds ratio, and percent outcome in the exposed group. 

* The *Unmatched Case-Control* input parameters: two-sided confidence level, power, of unexposed to exposed, percent outcome in the unexposed group, and odds ratio.

Positive aspects of *StatCalc* includes:
* The output lists sample sizes for the most popular confidence levels. There are seven options in the *Population Survey* module, six options in the *Cohort or Cross-Sectional* module, and six options in the *Unmatched Case-Control* module.


**Study type(s) this calculator would be suitable for:**

* Point in time surveys to estimate a proportion

* Humanitarian surveys

* Comparative surveys

* Study to test hypotheses


## 2. Review Checklist

1. What are the sampling assumptions made for the calculator?
All modules: Sampling without replacement.
The *Population Survey* module:
* Simple random sampling without replacement
* If systematic sampling, it must be representative. The population is in random order -- the listing of population units is not in some periodic or cyclical order.
* If elements are clustered, single-stage cluster sampling with clusters are of equal sizes


2. What measure of accuracy is used for sample size calculation? 
* The *Population Survey* module: Confidence interval, no mention of coefficient of variation
* The *Cohort or Cross-Section* module: Confidence level, power
* The *Unmatched Case-Control* module: Confidence level, power


3. If it includes continuous variables, how do they determine the variance of estimates since a distribution needs to be known or assumptions made about the distribution?
* Not applicable because the calculator does not include continuous variables. 


4. Is the calculator calculation correct?

The *Population Survey* module: for simple random sampling, the calculation of sample size to estimate population proportion is correct. For cluster sampling, we reconstructed the formula that the *Population Survey* module uses. The *Cohort or Cross-Sectional* module calculations are correct. The *Unmatched Case-Control* module calculations are correct. Section 4 provides details of how we verified the calculator calculation using R.


5. If clustering of sample units, is a design effect suggested? What about for double sampling for stratification?
The *Population Survey* module:
* Yes, a design effect is suggested for clustering of sample units
* The calculator only performs sampling plan calculation for single sampling.

The *Cohort or Cross-Sectional* module: not applicable

The *Unmatched Case-Control* module: not applicable


6. Is probability proportional to size, stratification, double sampling?
* No mention of probability proportional to size, stratification, or double sampling.


7. Accuracy of continuous variable estimate as well as that of proportion?
* The calculator only deals with proportion.


8. How does the calculator deal with the effect of non-response bias on accuracy?
* No mention of non-response.

## 3. Limitations

All modules:

* Measure of accuracy is limited to the eight default confidence levels, which users cannot customize.

The *Population Survey* module:

* It does not cover sample size calculation for surveys that involve stratified sampling.

* The labeling of input parameters "Clusters" and "Cluster Size" can confuse users. For example, the *Population Survey*'s window of  *StatCalc* provides an instruction: "For simple random sampling, leave design effect and clusters equal to 1". Users must look at the values under the output parameter "Cluster Size" to obtain the resulting sample size. That is, although labeled as "Cluster Size", its value is the sample size calculated by the *Population Survey* module. Additionally, in this simple random sampling case, if users are required to enter a value of 1 for the input parameter "Clusters", then the "Cluster Size" should equal to the value of the input parameter "Population size". But again, the value of "Cluster Size" is the resulting sample size instead. 

* The definition of the input parameter "Clusters" could be misunderstood. The *Population Survey* guideline defines input parameter "Clusters" as follows: "This input designates the number of clusters in the population survey." One possible interpretation of this definition is that the parameter "Clusters" is the number of clusters in the population. Another interpretation is that "Clusters" refers to the number of clusters in the sample. Following the former definition and assuming that clusters are of equal sizes, the value of "Cluster Size" should equal to the value of "Population size" divided by the value of "Clusters". However, the value of "Cluster Size" does not equal to the value of "Population size" divided by the value of "Clusters". Instead, "Cluster Size" is equal to "Total sample" divided by "Clusters". Thus, "Clusters" refers to the number of primary sampling units (psus) in the sample.


## 4. Verification of the Select Statistics calculator calculation

### 4.1 Population Survey

**Simple Random Sampling**

Formula to calculate sample size for estimating population proportion from simple random sampling is:

$$
n = \frac{N \times X}{(X + N - 1)}
$$

where,

$$
X = \frac{z_{\alpha/2}^2 \times p \times (1-p)}{\text{MOE}^2}
$$
and,

* $N$ : population size

* $z_{\alpha/2}$ : the critical value of the normal distribution at $\alpha/2$, where $\text{confidence level} = 1 - \alpha$

* $p$ : expected sample proportion

* $\text{MOE}$ : acceptable margin of error

The two equations above correspond to equations (4.3) and (4.2) in Cochran (1977), respectively, with the exception that the equations listed in this review document use normal distribution instead of t-distribution for the critical value.

```{r}
# create a function that calculates sample size based for SRS
compute_n_srs <- function(moe = 0.05, conf_level = 0.95, N = 100000, p = 0.5) {
 # critical value
 z <- qnorm((1-conf_level)/2, lower.tail = FALSE)
 # calculate X
 X <- (z^2 * p * (1-p)) / (moe^2)
 # calculate n
 n <-  (N * X) / (X + N - 1)
 n <- ceiling(n)
 return(n)
}
```


```{r}
# set input parameters: set 1
moe <- 0.05           # margin of error
conf_level <- 0.90    # confidence level
N <- 100000           # population size
p <- 0.5              # expected sample proportion

```

```{r}
# calculate recommended sample size
recommended_n <- compute_n_srs(moe = moe, conf_level = conf_level, N = N, p = p)
recommended_n
```

Given the input parameters of set 1, the *Population Survey* output for "Cluster Size" was $n$ = 270. Our verification resulted in $n$ = `r recommended_n`. The calculator correctly calculated $n$.

**Cluster Sampling**

To calculate the output parameter "Cluster Size", the *Population Survey* seems to use the following formula:
$$
\text{Cluster Size} = \frac{n_{srs} \times deff}{\text{Clusters}}
$$

where:
* $n_{srs}$ : sample size for simple random sampling using formulas in the earlier subsection

* $deff$ : design effect

* Clusters : the input parameter "Clusters". The *StatCalc* user guide explanation of "Clusters" is: "This input designates the number of clusters in the population survey". It is likely to mean the number of primary sampling units (psus) in the sample.

The numerator in the equation corresponds to total elements that would be needed, an approach introduced by Cornfield (1951) and noted in Cochran (1977) as well as in Lohr (2010).

With regard to the formula for "Total Sample", the user guide explains that "Total Sample" is "Cluster Size" multiplied by "Clusters":


$$
\text{Total Sample} = \text{Cluster Size} \times \text{Clusters}
$$

```{r}
# create a function to calculate cluster size
compute_cluster_size <- function(moe = 0.05, conf_level = 0.95, N = 100000, p = 0.5, deff = 1, clusters = 1) {
  n_srs <- compute_n_srs(moe = moe, conf_level = conf_level, N = N, p = p)
  cluster_size <- ceiling((n_srs * deff) / clusters)
  total_sample <- cluster_size * clusters
  print(paste0("Cluster Size : ", as.character(cluster_size)))
  print(paste0("Total Sample : ", as.character(total_sample)))
}
```

```{r}
# set input parameters: set 2
moe <- 0.05           # margin of error
conf_level <- 0.95    # confidence level
N <- 100000           # population size
p <- 0.5              # expected sample proportion
deff <- 1.5           # design effect
clusters <- 2
```


```{r}
# compute cluster size using input parameters set 2
compute_cluster_size(moe = moe, conf_level = conf_level, N = N, p = p, deff = deff, clusters = clusters)
```

Given the input parameters of set 2, the *Population Survey* output for "Cluster Size" was 288 and "Total Sample" was 576.

```{r}
# set input parameters: set 3
moe <- 0.05           # margin of error
conf_level <- 0.90    # confidence level
N <- 100000           # population size
p <- 0.5              # expected sample proportion
deff <- 1.5           # design effect
clusters <- 3
```

```{r}
# compute cluster size using input parameters set 3
compute_cluster_size(moe = moe, conf_level = conf_level, N = N, p = p, deff = deff, clusters = clusters)
```

Given the input parameters of set 2, the *Population Survey* output for "Cluster Size" was 135 and "Total Sample" was 405.

### 4.2 Unmatched Cohort and Cross Sectional

Formulas to calculate sample sizes as described in Kelsey et al. (1996) cf. Sullivan and Soe (2007):

$$
n_1 = \frac{(z_{\alpha/2} + z_{1-\beta})^2 \overline{p} \overline{q}(r+1)}{r(p_1 -1 p_2)^2}
$$
and

$$
n_2 = rn_1
$$
where:

* $n_1$ : number of exposed

* $n_2$ : number of unexposed

* $z_{\alpha/2}$ : standard normal deviate for two-tailed test based on alpha level (1 - confidence level)

* $z_{\beta}$ : standard normal deviate for one-tailed test based on beta level (1 - power)

* $r$ : ratio of exposed to unexposed

* $p_1$ : proportion of exposed with disease

* $p_2$ : proportion of unexposed with disease

* $\overline{p} = \frac{p_1 + r p_2}{r+ 1}$

* $\overline{q} = 1 - \overline{p}$

Formulas for sample sizes without the correction factor as described in Fleiss (1981) cf. Sullivan and Soe (2007):

$$
n_1 = \frac{\left[  z_{\alpha/2} \sqrt{(r+1)\overline{p}\overline{q}} + z_{1-\beta} \sqrt{rp_1q_1 + p_2q_2}   \right]^2}{r(p_1 - p_2)^2}
$$
and

$$
n_2 = rn_1
$$

Formulas for sample sizes with the correction factor as described in Fleiss (1981) cf. Kasiulevičius et al. (2006):

$$
n_{1cc} = \frac{n_1}{4} \left[ 1 + \sqrt{1 + \frac{2(r+1)}{n_1r|p_2 - p_1|}} \right]^2
$$

and

$$
n_{2cc} = rn_{1cc}
$$
Formula for an odds ratio when the input provided is proportion of exposed:
$$
OR = \frac{p_2(1-p_1)}{p_1(1 - p_2)}
$$


Formula for proportion of exposed when the input provided is an odds ratio:
$$
p_1 = \frac{p_2 OR}{1 + p_2 (OR -1)}
$$

Formula for proportion of exposed when the input provided is a risk ratio (RR):

$$
p_1 = p_2 RR
$$

**Verify sample sizes using Kesley formulas**

```{r}
# create function to calculate sample sizes using Kelsey et. al (1996)
# input provided : an odds ratio

compute_n_kelsey_or <- function(confidence_level = 0.95, power = 0.8, OR, r = 1, p2) {
  alpha <- 1 - confidence_level
  z_alpha_2tails <- qnorm(alpha/2, lower.tail = FALSE)
  z_power <- qnorm(power, lower.tail = TRUE)
  p1 <- (p2 * OR)/(1 + p2*(OR - 1))
  p_bar <- (p1 + r*p2) / (r + 1)
  n1 <- ((z_alpha_2tails + z_power)^2 * p_bar * (1 - p_bar) * (r + 1) ) / (r * (p1 - p2)^2)
  n1 <- ceiling(n1)
  n2 <- r * n1
  print(paste0("proportion outcome in the exposed group (p1) : ", as.character(round(p1,2))))
  print(paste0("sample size of exposed group (n1) : ", as.character(n1)))
  print(paste0("sample size of unexposed group (n2) : ", as.character(n2)))
}
```

```{r}
# set input parameters: set 4
confidence_level <- 0.95
power <- 0.8
r <- 1
p2 <- 0.075
OR <- 2
```

```{r}
# calculate n1 using input parameters set 4
compute_n_kelsey_or(confidence_level = confidence_level, power = power, OR = OR, r = r, p2 = p2)
```

Given identical input parameters as in set 4, the *Cohort or Cross-Sectional* module also generated $p_1$ = 14%, and sample sizes using Kelsey method: $n_1$ = 361, and $n_2$ = 361.

**Verify sample sizes using Fleiss formulas (without correction)**

```{r}
# create function to calculate sample sizes using Fleiss (1981) without the correction factor
# input provided : an odds ratio

compute_n_fleiss_or <- function(confidence_level = 0.95, power = 0.8, OR, r = 1, p2) {
  # prepare inputs
  alpha <- 1 - confidence_level
  z_alpha_2tails <- qnorm(alpha/2, lower.tail = FALSE)
  z_power <- qnorm(power, lower.tail = TRUE)
  p1 <- (p2 * OR)/(1 + p2*(OR - 1))
  p_bar <- (p1 + r*p2) / (r + 1)
  q1 <- 1 - p1
  q2 <- 1 - p2
  q_bar <- 1 - p_bar
  
  # calculate sample sizes
  numerator_n1 <- ( (z_alpha_2tails * sqrt((r+1)*p_bar*q_bar)) + (z_power * sqrt(r*p1*q1 + p2*q2)) )^2
  denominator_n1 <- (r * (p1 - p2)^2)
  n1 <-  numerator_n1 / denominator_n1 
  n1 <- ceiling(n1)
  n2 <- r * n1
  print(paste0("proportion outcome in the exposed group (p1) : ", as.character(round(p1,2))))
  print(paste0("sample size of exposed group (n1) : ", as.character(n1)))
  print(paste0("sample size of unexposed group (n2) : ", as.character(n2)))
}
```

```{r}
# calculate n1 using input parameters set 4
compute_n_fleiss_or(confidence_level = confidence_level, power = power, OR = OR, r = r, p2 = p2)
```

Given identical input parameters as in set 4, the *Cohort or Cross-Sectional* module also generated $p_1$ = 14%, and sample sizes using Fleiss method without correction: $n_1$ = 360, and $n_2$ = 360.

**Verify sample sizes using Fleiss formulas (with correction)**

```{r}
# create function to calculate sample sizes using Fleiss (1981) with the correction factor
# input provided : an odds ratio

compute_n_fleiss_cc_or <- function(confidence_level = 0.95, power = 0.8, OR, r = 1, p2) {
  # prepare inputs
  alpha <- 1 - confidence_level
  z_alpha_2tails <- qnorm(alpha/2, lower.tail = FALSE)
  z_power <- qnorm(power, lower.tail = TRUE)
  p1 <- (p2 * OR)/(1 + p2*(OR - 1))
  p_bar <- (p1 + r*p2) / (r + 1)
  q1 <- 1 - p1
  q2 <- 1 - p2
  q_bar <- 1 - p_bar
  
  # calculate sample sizes - without correction
  numerator_n1 <- ( (z_alpha_2tails * sqrt((r+1)*p_bar*q_bar)) + (z_power * sqrt(r*p1*q1 + p2*q2)) )^2
  denominator_n1 <- (r * (p1 - p2)^2)
  n1 <-  numerator_n1 / denominator_n1 
  n1 <- ceiling(n1)
  
  # calculate sample sizes - with correction
  n1cc <- (n1/4) * (1 + sqrt(1 + 2*(r+1) / (n1*r*abs(p2-p1))))^2
  n1cc <- ceiling(n1cc)
  n2cc <- r * n1cc
  print(paste0("proportion outcome in the exposed group (p1) : ", as.character(round(p1,2))))
  print(paste0("sample size of exposed group (n1cc) : ", as.character(n1cc)))
  print(paste0("sample size of unexposed group (n2cc) : ", as.character(n2cc)))
}
```


```{r}
# calculate n1 using input parameters set 4
compute_n_fleiss_cc_or(confidence_level = confidence_level, power = power, OR = OR, r = r, p2 = p2)
```

Given identical input parameters as in set 4, the *Cohort or Cross-Sectional* module also generated $p_1$ = 14%, and sample sizes using Fleiss method with correction: $n_1$ = 391, and $n_2$ = 391.

### 4.3 Unmatched Case Control

The *Unmatched Case Control* module uses the same formulas as the *Cohort or Cross-Sectional* module to calculate sample size. Using input parameters set 4, we obtain outputs as noted above. The *Unmatched Case Control* generated identical outputs. That is, $p_1$ = 14%, and sample sizes using Kelsey method: $n_1$ = 361, and $n_2$ = 361. Sample sizes using Fleiss method without correction: $n_1$ = 360, and $n_2$ = 360. And sample sizes using Fleiss method with correction: $n_1$ = 391, and $n_2$ = 391.


## 5. Reference

Cochran, W. G. (1977). *Sampling techniques*. John Wiley & Sons.

Cornfield, J. (1951). The determination of sample size. *American Journal of Public Health and the Nations Health*, 41(6), 654-661.

Fleiss J.L. (1981). *Statistical Methods for Rates and Proportions*. John Wiley & Sons. 

Kasiulevičius, V., Šapoka, V., & Filipavičiūtė, R. (2006). Sample size calculation in epidemiological studies. *Gerontologija*, 7(4), 225-231.

Kelsey, J. L., Kelsey, W. E., Whittemore, A. S., Evans, A. S., & Thompson, W. D. (1996). *Methods in Observational Epidemiology*. Oxford University Press.

Lohr, S. L. (2010). *Sampling: design and analysis*. Chapman and Hall/CRC.

Sullivan, K. M., Soe, M. M. (2007, February). *Sample Size for a Cross-Sectional, Cohort, or Clinical Trial Studies*. http://www.openepi.com/PDFDocs/SSCohortDoc.pdf.

Sullivan, K. M., Soe, M. M. (2007, February). *Sample Size for an Unmatched Case-Control Study*. http://www.openepi.com/PDFDocs/SSCCDoc.pdf.





