---
title: "Review of LQAS Sampling Plan Calculator"
author: "Team 2"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Source link: [Brixton Health's sampling plan calculator](http://www.brixtonhealth.com/hyperLQAS.findD.html)

## 1. Summary

The *brixtonhealth* calculator calculates a decision rule, actual type 1 error, and actual type 2 error given the generated decision rule for a lot quality assurance sampling (LQAS). To obtain these three outputs, user provides as input: the population size, the upper proportion threshold to construct the null hypothesis, the lower proportion threshold, sample size, maximum tolerable type 1 error, and maximum tolerable type 2 error. *brixtonhealth* calculation is based on hypergeometric distribution.

Study type(s) this calculator would be suitable for: 
* Point in time surveys to estimate a proportion
* Humanitarian surveys
* Comparative surveys


## 2. Review checklist

1. What are the sampling assumptions  made for the calculator? 
* simple random sampling without replacement
* one-tail test


2. What measure of accuracy is used for sample size calculation? 
* no mention of confidence interval
* no mention of coefficient of variation
* other: type I error, $\alpha$, and type II error, $\beta$.


3. If it includes continuous variables how do they determine variance of estimates since a distribution needs to be known or assumptions made about the distribution?
* not applicable because the calculator does not include continuous variables. 


4. Is the calculator calculation correct?
The calculator outputs are: decision rule, actual $\alpha$ error, and actual $\beta$ error. The calculation of actual $\alpha$ error and actual $\beta$ error that *brixtonhealth* calculator generates are correct. The decision rule that *brixtonhealth* generates fulfills the conditions of lower than the maximum tolerable type 1 error and the maximum tolerable type 2 error that user specified. Section 3 provides details of how we checked the calculator calculation.


5. If clustering of sample units, is a design effect suggested? What about for double sampling for stratification?
* The sampling unit is not cluster.
* No mention of design effect.
* The calculator only performs sampling plan calculation for single sampling.


6. Is probability proportional to size, stratification, double sampling?
* LQAS is identical to stratified sampling in which the samples are too small to provide acceptably narrow confidence intervals for estimates (Lemeshow and Stroh, 1989).


7. Accuracy of continuous variable estimate as well as that of proportion?
* The calculator only deals with proportion.


8. How does calculator deal with effect of non-response bias on accuracy?
* No mention of non-response.



## 2. Checking the calculator calculation

**Summary:** the calculation of actual $\alpha$ and actual $\beta$ that *brixtonhealth* calculator generates are correct and identical to the approach described in Lemeshow and Stroh (1989). Its resulting decision rule is slightly different than the decision rule that we calculated using Lemeshow and Stroh (1989) as reference. The *brixtonhealth* calculator uses hypergeometric distribution rather than normal approximation of hypergeometric distribution.

The *brixtonhealth* calculator uses the following default specifications:

* Population size, $N = 600$

* Upper threshold, $P_0 = 0.8$

* Lower threshold, $P_a = 0.5$

* Sample size, $n = 50$

* Maximum tolerable $\alpha$ error = 0.1

* Maximum tolerable $\beta$ error = 0.1


```{r}
# specifications
N <- 600      # population size
n <- 50       # sample size
p_0 <- 0.8    # upper threshold
p_a <- 0.5    # lower threshold
alpha <- 0.1
beta <- 0.1

```


**Check actual alpha and actual beta: ** 

Using the above specifications, the *brixtonhealth* calculator generates decision rule $d^*$ = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057.

With $d^*$ = 33, we check the calculation of $\alpha$ and $\beta$:  

```{r}
type1_error_hyper <- phyper(q=33, m = N*p_0, n = (N-N*p_0), k = n)
type1_error_hyper
```

```{r}
power_of_test <- phyper(q=33, m = N*p_a, n = (N-N*p_a), k = n)
type2_error_hyper <- 1 - power_of_test
type2_error_hyper
```

We expect to get identical results as *brixtonhealth* calculator generates. That is, type 1 error = 0.0110 and type 2 error = 0.0057. We get `r round(type1_error_hyper,4)` and `r round(type2_error_hyper,4)` for type 1 error and type 2 error, respectively. Our calculation of actual $\alpha$ and actual $\beta$ are identical to those of *brixtonhealth* calculator. 

**Check decision rule:**

```{r}
# create a function to find d*, decision rule based on alpha
find_dstar <- function(N, p, n, alpha){
  dstar = 0
  for (i in seq(from = 0, to = n)) {
    type_1_error <- phyper(q=i, m = N*p, n = (N-N*p), k = n)
    if (type_1_error <= alpha) {
      dstar = i
    }
  }
  return(dstar)
}
```

```{r}
d_star <- find_dstar(N=N, p=p_0, n=n, alpha=alpha)
d_star
```

```{r}
# recheck the resulting type 1 error
type1_error_hyper <- phyper(q=d_star, m = N*p_0, n = (N-N*p_0), k = n)
type1_error_hyper
```

Resulting actual $\alpha$ is `r round(type1_error_hyper,4)`, lower than the maximum allowable $\alpha$ of 0.1.

```{r}
# check type 2 error using d_star
power_of_test <- phyper(q=d_star, m = N*p_a, n = (N-N*p_a), k = n)
type2_error_hyper <- 1 - power_of_test
type2_error_hyper
```

Resulting actual $\beta$ is `r round(type2_error_hyper,6)`, lower than the maximum tolerable $\beta$ of 0.1.

We expect to get decision rule of 33, as generated by *brixtonhealth* calculator. But, our calculation thus far yields decision rule of 35. 

Next, we explore the possibility of considering maximum tolerable $\beta$ in generating decision rule.

```{r}
# find d* based on beta
d_star_beta <- find_dstar(N=N, p=p_a, n=n, alpha=beta)
d_star_beta

```

```{r}
# check type 1 error based on d_star_beta
phyper(q=d_star_beta, m = N*p_0, n = (N-N*p_0), k = n)

```

```{r}
# check type 2 error based on d_star_beta
power_tmp <- phyper(q=d_star_beta, m = N*p_a, n = (N-N*p_a), k = n)
type_2_error <- 1-power_tmp
type_2_error
```

Considering $\beta$ does not lead to decision rule of 33, as generated by *brexithealth* calculator.

Next we consider decision rule based on $\alpha$ as a starting point, and progressively reduces it until it reach the maximum tolerable $\beta$

```{r}
type1_error_hyper
```

```{r}
type2_error_hyper
```

```{r}
# type 1 error using alternative decision rule
phyper(q=d_star-6, m = N*p_0, n = (N-N*p_0), k = n)
```
```{r}
# type 2 error using alternative decision rule
1 - phyper(q=d_star-6, m = N*p_a, n = (N-N*p_a), k = n)
```

```{r}
d_star-6
```

This approach results in decision rule of 29.

The decision rule that *brixtonhealth* produces is within the range of our calculated decision rules. Additionally, the decision rule results in type 1 error and type 2 error that are lower than the maximum allowable type 1 error and type 2 error that user specified. Therefore, we conclude that the calculator correctly calculates the decision rule. 

```{r}
find_dstar(N=N, p=p_0, n=n, alpha=0.02)
```

```{r}
phyper(q=33, m = N*p_0, n = (N-N*p_0), k = n)
```

```{r}
find_dstar(N=N, p=p_0, n=40, alpha=0.02)
```
```{r}
phyper(q=26, m = N*p_0, n = (N-N*p_0), k = 40)
```

```{r}
find_dstar(N=N, p=p_0, n=25, alpha=0.05)
```

```{r}
phyper(q=16, m = N*p_0, n = (N-N*p_0), k = 25)
```

```{r}
find_dstar(N=N, p=p_0, n=15, alpha=0.1)
```


**Input validation check**

*check 1:* input: sample size = none (left empty) and other inputs are their default values, outputs: decision rule = 0, actual $\alpha$ error = 1, and actual $\beta$ error = 0. Calculator provides outputs even when sample size is missing.

*check 2:* input: sample size = 0 and other inputs are their default values, outputs: warning "Sample size is too small".

*check 3:* input: maximum tolerable $\alpha$ error = none (left empty) and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. Although the value of maximum tolerable $\alpha$ error is missing (not specified), the calculator provides outputs that are identical to outputs if maximum tolerable $\alpha$ error = 0.1.

*check 4:* input: maximum tolerable $\alpha$ error = 0 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\alpha$ error = 0.1.

*check 5:* input: maximum tolerable $\alpha$ error = 0.05 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\alpha$ error = 0.1.

*check 6:* input: maximum tolerable $\alpha$ error = 0.2 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\alpha$ error = 0.1.

*check 7:* input: maximum tolerable $\beta$ error = none (left empty) and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. Although the value of maximum tolerable $\alpha$ error is missing (not specified), the calculator provides outputs that are identical to outputs if maximum tolerable $\beta$ error = 0.1.

*check 8:* input: maximum tolerable $\beta$ error = 0 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\beta$ error = 0.1.

*check 9:* input: maximum tolerable $\beta$ error = 0.05 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\beta$ error = 0.1.

*check 10:* input: maximum tolerable $\beta$ error = 0.2 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\beta$ error = 0.1.

*check 11:* input: sample size = 40 and other inputs are their default values, outputs: decision rule = 26, actual $\alpha$ error = 0.0160, and actual $\beta$ error = 0.0161. Given sample size = 40 and decision rule = 26, the calculator generates correct actual $\alpha$ error and actual $\beta$ error.



## Limitation

* The calculator 


### 2.2 Check brixtonhealth calculation using Lemeshow & Taber (1991) as reference

Lemeshow & Taber (1991) uses normal approximation to hypergeometric distribution. 

Hypergeometric distribution is represented by the following formula:
$$
\mathrm{P}(d \le d^*) = \sum_{d=0}^{d^*} \frac{\binom{NP_0}{d} \binom{N(1-P_0)}{n-d}}{\binom{N}{n}}
$$
The mean and variance of hypergeometric distribution with sample size $n$ and the hypothesized proportion $P_0$ are:

$$
\mathrm{mean}(d) = nP_0
$$
and
$$
\mathrm{var}(d) = nP_0(1-P_0)\left( \frac{N-n}{N-1} \right)
$$

If the calculator uses normal approximation to the hypergeometric distribution, then $d$ would be calculated as:
$$
d^* = nP_0 - z_{1-\alpha} \sqrt{nP_0(1-P_0)\left( \frac{N-n}{N-1}\right)}  
$$
Because $P_a$ and $\beta$ are also specified in the calculator, 

$$
d^* = nP_0 - z_{1-\beta} \sqrt{nP_a(1-P_a)\left( \frac{N-n}{N-1}\right)}  
$$

Setting the above two equations for $d$ equal to each other, and solving for $n$:
$$
n = \left(\frac{z_{1-\alpha} \sqrt{P_0(1-P_0)} + z_{1-\beta} \sqrt{P_a(1-P_a)}}{(P_0 - P_a)}  \right)^2
$$


```{r}
N <- 1000000 # set population size so that N >> n

z_0 <- qnorm(alpha, lower.tail = FALSE)
paste0("z_0: ", round(z_0,3))
z_a <- qnorm(beta, lower.tail = FALSE)
paste0("z_a: ", round(z_a,3))

# calculate n that makes the two d* equations equals
numer <- z_0*sqrt(p_0*(1-p_0)) + z_a*sqrt(p_a*(1-p_a))
denom <- p_0 - p_a

n_calc <- ceiling((numer / denom)^2)
d_star <- n_calc*p_0 - (z_0 * sqrt(n_calc*p_0*(1-p_0)))

paste0("n: ", n_calc)
paste0("d_star: ", ceiling(d_star))
```

We expect to get identical results as *brexithealth* calculator generates. That is, decision rule = 33. However, we get decision rule =  `r ceiling(d_star)`.

It could be that we do not get decision rule of 33 because we use maximum tolerable $\alpha$ and $\beta$ instead of actual $\alpha$ and actual $\beta$. We recalculate decision rule using actual $\alpha$ = 0.0110 and actual $\beta$ = 0.0057.

```{r}
actual_alpha <- 0.0110
actual_beta <- 0.0057

z_0_act <- qnorm(actual_alpha, lower.tail = FALSE)
paste0("z_0: ", round(z_0_act,3))
z_a_act <- qnorm(actual_beta, lower.tail = FALSE)
paste0("z_a: ", round(z_a_act,3))

# calculate n that makes the two d* equations equals
numer <- z_0*sqrt(p_0*(1-p_0)) + z_a*sqrt(p_a*(1-p_a))
denom <- p_0 - p_a

n_calc <- ceiling((numer / denom)^2)
d_star <- n_calc*p_0 - (z_0 * sqrt(n_calc*p_0*(1-p_0)))

paste0("n: ", n_calc)
paste0("d_star: ", ceiling(d_star))

```

The decision rule using actual $\alpha$ and actual $\beta$ is `r ceiling(d_star)`. It remains different from decision rule generated by *brexithealth*.

Conclusion: the *brexithealth* calculator does not seem to use normal approximation to hypergeometric distribution.


