---
title: "Review of LQAS Sampling Plan Calculator"
subtitle: "SQUEAC Calculator"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Source: `SQUEAC.Setup.exe`

## 1. Summary

The *squeac* calculator calculates sample sizes and decision rules (decision thresholds) for a LQAS. Its input parameters are upper standard (upper threshold proportion), lower standard (lower threshold proportion), lower error level (type 1 error), and upper error level (type 2 error). Given these input parameters, the calculator generates a list of pairs of sample size and its decision rule, and their associated actual lower error level (actual type 1 error) and actual upper error level (actual type 2 error).

Study type(s) this calculator would be suitable for:

* Point in time surveys to estimate a proportion

* Humanitarian surveys

* Comparative surveys


## 2. Review Checklist

1. What are the sampling assumptions  made for the calculator? 
* Simple random sampling without replacement
* Population size is very large relative to the sample size such that computing sample sizes and their errors (type 1 and type 2) can be accomplished using the normal distribution and the binomial distribution, respectively. 
* One-sided test


2. What measure of accuracy is used for sample size calculation? 
* No mention of confidence interval
* No mention of coefficient of variation
* Other: upper error level (type 1 error) and lower error level (type 2 error).


3. If it includes continuous variables how do they determine variance of estimates since a distribution needs to be known or assumptions made about the distribution?
* Not applicable because the calculator does not include continuous variables. 


4. Is the calculator calculation correct?

For each pair of sample size and decision rule that the calculator generates, its actual upper level error (type 1 error) and actual lower level error (type 2 error) are calculated correctly. For the calculation of sample sizes and decision rules, the *squeac* calculator outputs are very similar to our our calculation. Section 4 provides details of how we verified the calculator calculation.


5. If clustering of sample units, is a design effect suggested? What about for double sampling for stratification?
* The sampling unit is not cluster.
* No mention of design effect.
* The calculator only performs sampling plan calculation for single sampling.


6. Is probability proportional to size, stratification, double sampling?
* LQAS is identical to stratified sampling in which the samples are too small to provide acceptably narrow confidence intervals for estimates (Lemeshow and Stroh, 1989).


7. Accuracy of continuous variable estimate as well as that of proportion?
* The calculator only deals with proportion.


8. How does calculator deal with effect of non-response bias on accuracy?
* No mention of non-response.


## 3. Limitations

* The calculator is limited to LQAS single sampling plans. It does not provide guidance for LQAS double sampling plans.

* It is applicability is limited to situations where the population size is very large relative to the sample size. It is not suitable for situations where sample is drawn without replacement from a small population. 

* The power achieved from the sampling is not directly available for users. To obtain the power achieved from the sampling plan, users must conduct an additional step of calculation. 


## 4. Verification of the SQUEAC calculator calculation

### 4.1 Verify calculation: input parameters set 1

To check the calculation of the *squeac* calculator, we use input parameters as follow: lower standard = 50%, upper standard = 80%, upper error level = 10%, and lower error level = 10%. These input parameters are similar to those in Lemeshow and Taber (1991).

```{r}
# set input parameters
alpha <- 0.05 # upper error level
beta <- 0.2 # lower error level
p_0 <- 0.2    # upper standard
p_a <- 0.1    # lower standard

```

If the calculator uses normal approximation to the hypergeometric distribution, following Lemeshow and Taber (1991), the formula for sample size is: 

Sample size:
$$
n = \left(\frac{z_{1-\alpha} \sqrt{P_0(1-P_0)} + z_{1-\beta} \sqrt{P_a(1-P_a)}}{(P_0 - P_a)}  \right)^2
$$

```{r}
z_0 <- qnorm(alpha, lower.tail = FALSE)
paste0("z_0: ", round(z_0,3))
z_a <- qnorm(beta, lower.tail = FALSE)
paste0("z_a: ", round(z_a,3))
```

```{r}
# calculate n
numer <- z_0*sqrt(p_0*(1-p_0)) + z_a*sqrt(p_a*(1-p_a))
denom <- p_0 - p_a

n_calc <- ceiling((numer / denom)^2)
d_star <- n_calc*p_0 - (z_0 * sqrt(n_calc*p_0*(1-p_0)))

paste0("n: ", n_calc)
paste0("d_star: ", ceiling(d_star))
```

The *squeac* calculator generates outputs that start from n = 81 and d = 10. Very similar.

```{r}
# calculate P(d<=11)
pbinom(11, size = 83, prob = 0.2)
```

Higher than the specified $\alpha$, upper error level = 0.05


**check upper and lower given n = 81 and d = 10**

```{r}
# calculate type 1 error using calculator output: n = 81, d = 10.
# other input parameter: p_0 (upper standard) = 0.2
type_1_error <- pbinom(q=10, size = 81, prob = 0.2)
type_1_error # actual upper error level
```
The *squeac* calculator generates upper error = 5%. Thus, *squeac* output for upper error matches with our calculation of type 1 error = `r round(type_1_error, 3)`.

```{r}
# check type 2 error / "lower error" output, using calculator output: n = 81, d 10
# other input parameter: p_a (lower standard) = 0.1
1-pbinom(q=10, size = 81, prob = 0.1) # actual lower error level
```
The resulting calculation is similar to the output of the *squeac* calculator = 18%.


**check upper and lower given n = 82 and d = 10**
```{r}
# calculate type 1 error using calculator output: n = 82, d = 10.
# other input parameter: p_0 (upper standard) = 0.2
type_1_error <- pbinom(q=10, size = 82, prob = 0.2)
type_1_error # actual upper error level
```

The *squeac* calculator output, upper = 5%. very close. 


```{r}
# check type 2 error / "lower error" output, using calculator output: n = 82, d 10
# other input parameter: p_a (lower standard) = 0.1
1-pbinom(q=10, size = 82, prob = 0.1) # actual lower error level
```

The *squeac* calculator output, lower = 19%. correct.


### 4.2 Verify calculation: input parameters set 2

Check with different input parameters

```{r}
# set different input parameters
alpha <- 0.01 # upper error level
beta <- 0.2 # lower error level
p_0 <- 0.8    # upper standard
p_a <- 0.5    # lower standard
```

```{r}
z_0 <- qnorm(alpha, lower.tail = FALSE)
paste0("z_0: ", round(z_0,3))
z_a <- qnorm(beta, lower.tail = FALSE)
paste0("z_a: ", round(z_a,3))
```

```{r}
# calculate n
numer <- z_0*sqrt(p_0*(1-p_0)) + z_a*sqrt(p_a*(1-p_a))
denom <- p_0 - p_a

n_calc <- ceiling((numer / denom)^2)
d_star <- n_calc*p_0 - (z_0 * sqrt(n_calc*p_0*(1-p_0)))

paste0("n: ", n_calc)
paste0("d_star: ", ceiling(d_star))
```

calculator outputs start from n = 21 and d = 12. very close.

```{r}
# calculate P(d<= 13)
pbinom(q=13, size = 21, prob = p_0)
```

Higher than the specified $\alpha$, upper error level = 0.01

**check upper and lower given n = 21 and d = 12**

```{r}
# calculate type 1 error: n = 21, d = 13.
# other input parameter: p_0 (upper standard) = p_0 = 0.8
type_1_error <- pbinom(q=12, size = 21, prob = p_0)
type_1_error # actual upper error level
```

The *squeac* calculator output = 1%. correct.

```{r}
# check type 2 error / "lower error" output, using calculator output: n = 81, d 10
# other input parameter: p_a (lower standard) = 0.5
1-pbinom(q=12, size = 21, prob = p_a) # actual lower error level
```

The *squeac* calculator ouput = 19%. correct.

### 4.3 Input validation check

*check 1*: lower standard = none (left missing), upper standard = 10%, and other inputs are their default values, outputs: warning "You must specify a number (%) for the lower level".

*check 2*: upper standard = none (left missing), lower standard = 5%, and other inputs are their default values, outputs: warning "You must specify a number (%) for the upper level".

*check 3*: lower standard = 10%, upper standard = 5%, and other inputs are their default values, outputs: warning "Upper level may not be less than lower level".

*check 4*: lower standard = 10%, upper standard = 5%, upper error = none (left missing), and lower error is at its default value, outputs: "No plans found!".

*check 5*: lower standard = 10%, upper standard = 5%, lower error = none (left missing), and upper error is at its default value, outputs: "No plans found!".

