---
title: "Review of LQAS Sampling Plan Calculator"
subtitle: "Brixton Health Calculator"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Source link: [Brixton Health's sampling plan calculator](http://www.brixtonhealth.com/hyperLQAS.findD.html)

## 1. Summary

The *brixtonhealth* calculator calculates decision rule (decision threshold), actual type 1 error, and actual type 2 error. To obtain these three outputs, a user provides as input parameters: the population size, the upper proportion threshold to construct the null hypothesis, the lower proportion threshold, sample size, maximum tolerable type 1 error, and maximum tolerable type 2 error. For the actual type 1 error and actual type 2 error, these numbers are calculated based on the decision rule that the *brexihealth* generates and the sample size, population size, upper threshold, and lower threshold that users specified. The *brixtonhealth* calculation is based on the hypergeometric distribution.

Study type(s) this calculator would be suitable for:

* Point in time surveys to estimate a proportion

* Humanitarian surveys

* Comparative surveys



## 2. Review checklist

1. What are the sampling assumptions  made for the calculator? 
* Simple random sampling without replacement
* One-sided test


2. What measure of accuracy is used for sample size calculation? 
* No mention of confidence interval
* No mention of coefficient of variation
* Other: type I error, $\alpha$, and type II error, $\beta$.


3. If it includes continuous variables how do they determine variance of estimates since a distribution needs to be known or assumptions made about the distribution?
* Not applicable because the calculator does not include continuous variables. 


4. Is the calculator calculation correct?

The calculation of actual $\alpha$ error and actual $\beta$ error that the *brixtonhealth* calculator generates are correct for the decision rule that the calculator produces. The decision rule that the *brixtonhealth* generates fulfills the conditions of lower than the maximum tolerable type 1 error and the maximum tolerable type 2 error that user specified. Section 4 provides details of how we verified the calculator calculation.


5. If clustering of sample units, is a design effect suggested? What about for double sampling for stratification?
* The sampling unit is not cluster.
* No mention of design effect.
* The calculator only performs sampling plan calculation for single sampling.


6. Is probability proportional to size, stratification, double sampling?
* LQAS is identical to stratified sampling in which the samples are too small to provide acceptably narrow confidence intervals for estimates (Lemeshow and Stroh, 1989).


7. Accuracy of continuous variable estimate as well as that of proportion?
* The calculator only deals with proportion.


8. How does calculator deal with effect of non-response bias on accuracy?
* No mention of non-response.



## 3. Limitations

* The calculator is limited to LQAS single sampling plans. It does not provide guidance for LQAS double sampling plans.

* It does not calculate sample size. Users are expected to provide the sample size as part of the calculator input. 

* The calculator generates outputs even if some of the inputs are missing. 

* It is not always clear how the calculator utilizes the maximum tolerable $\alpha$ error and the maximum tolerable $\beta$ error that users specify. The calculator' outputs may remain the same even when the value of maximum tolerable $\alpha$ error or the maximum tolerable $\beta$ error varies. For example, see check #4, #5, #6 in the *Section 4: Input validation check* section.

* The power achieved from the sampling is not directly available to users. To obtain the power achieved from the sampling plan, users must conduct an additional step of calculation. 

* The calculator is limited to calculating LQAS sampling plan based on hypergeometric distribution. It does not cover LQAS sampling plans that are based on binomial approximation to the hypergeometric distribution, or normal approximation to the hypergeometric distribution.


## 4. Verification of the Brixton Health calculator calculation

**Summary:** the calculation of actual $\alpha$ and actual $\beta$ that the *brixtonhealth* calculator generates are correct and identical to the approach described in Lemeshow and Stroh (1989). Its resulting decision rule is slightly different than the decision rule that we calculated using Lemeshow and Stroh (1989) as reference. The *brixtonhealth* calculator uses hypergeometric distribution rather than normal approximation to the hypergeometric distribution.

Hypergeometric distribution is represented by the following formula:
$$
\mathrm{P}(d \le d^*) = \sum_{d=0}^{d^*} \frac{\binom{NP_0}{d} \binom{N(1-P_0)}{n-d}}{\binom{N}{n}}
$$

where $d$ is the number of "defect", such as unvaccinated people or HIV seropositive subjects; and $d^*$ is the decision rule (decision threshold).

The *brixtonhealth* calculator uses the following default specifications:

* Population size, $N = 600$

* Upper threshold, $P_0 = 0.8$

* Lower threshold, $P_a = 0.5$

* Sample size, $n = 50$

* Maximum tolerable $\alpha$ error = 0.1

* Maximum tolerable $\beta$ error = 0.1


```{r}
# specifications
N <- 600      # population size
n <- 50       # sample size
p_0 <- 0.8    # upper threshold
p_a <- 0.5    # lower threshold
alpha <- 0.1
beta <- 0.1

```


**Check actual alpha and actual beta: ** 

Using the above specifications, the *brixtonhealth* calculator generates decision rule $d^*$ = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057.

With $d^*$ = 33, we check the calculation of $\alpha$ and $\beta$:  

```{r}
type1_error_hyper <- phyper(q=33, m = N*p_0, n = (N-N*p_0), k = n)
type1_error_hyper
```

```{r}
power_of_test <- phyper(q=33, m = N*p_a, n = (N-N*p_a), k = n)
type2_error_hyper <- 1 - power_of_test
type2_error_hyper
```

We expect to get identical results as the *brixtonhealth* calculator generates. That is, type 1 error = 0.0110 and type 2 error = 0.0057. We get `r round(type1_error_hyper,4)` and `r round(type2_error_hyper,4)` for type 1 error and type 2 error, respectively. Our calculation of actual $\alpha$ and actual $\beta$ are identical to those of the *brixtonhealth* calculator. 

**Check decision rule:**

```{r}
# create a function to find d*, decision rule based on alpha
find_dstar <- function(N, p, n, alpha){
  dstar = 0
  for (i in seq(from = 0, to = n)) {
    type_1_error <- phyper(q=i, m = N*p, n = (N-N*p), k = n)
    if (type_1_error <= alpha) {
      dstar = i
    }
  }
  return(dstar)
}
```

```{r}
d_star <- find_dstar(N=N, p=p_0, n=n, alpha=alpha)
d_star
```

```{r}
# recheck the resulting type 1 error
type1_error_hyper <- phyper(q=d_star, m = N*p_0, n = (N-N*p_0), k = n)
type1_error_hyper
```

Resulting actual $\alpha$ is `r round(type1_error_hyper,4)`, lower than the maximum allowable $\alpha$ of 0.1.

```{r}
# check type 2 error using d_star
power_of_test <- phyper(q=d_star, m = N*p_a, n = (N-N*p_a), k = n)
type2_error_hyper <- 1 - power_of_test
type2_error_hyper
```

Resulting actual $\beta$ is `r round(type2_error_hyper,6)`, lower than the maximum tolerable $\beta$ of 0.1.

We expect to get decision rule of 33, as generated by the *brixtonhealth* calculator. But, our calculation thus far yields decision rule of 35. 

Next, we explore the possibility of considering maximum tolerable $\beta$ in generating decision rule.

```{r}
# find d* based on beta
d_star_beta <- find_dstar(N=N, p=p_a, n=n, alpha=beta)
d_star_beta

```

```{r}
# check type 1 error based on d_star_beta
phyper(q=d_star_beta, m = N*p_0, n = (N-N*p_0), k = n)

```

```{r}
# check type 2 error based on d_star_beta
power_tmp <- phyper(q=d_star_beta, m = N*p_a, n = (N-N*p_a), k = n)
type_2_error <- 1-power_tmp
type_2_error
```

Considering $\beta$ does not lead to decision rule of 33, as generated by the *brexithealth* calculator.

Next we consider decision rule based on $\alpha$ as a starting point, and progressively reduces it until it reach the maximum tolerable $\beta$.

```{r}
# type 1 error using alternative (reduced) decision rule
phyper(q=d_star-6, m = N*p_0, n = (N-N*p_0), k = n)
```
```{r}
# type 2 error using alternative (reduced) decision rule
1 - phyper(q=d_star-6, m = N*p_a, n = (N-N*p_a), k = n)
```

```{r}
d_star-6
```

This approach results in decision rule of 29.

The decision rule that the *brixtonhealth* produces is within the range of our calculated decision rules. Additionally, the decision rule results in type 1 error and type 2 error that are lower than the maximum allowable type 1 error and type 2 error that user specified. Therefore, we concluded that the calculator correctly calculates the decision rule. 

We also checked whether the calculator uses normal approximation to hypergeometric distribution, as explained in Lemeshow and Taber (1991). We follow formulas in Lemeshow and Taber (1991) to calculate sample size and decision rule. 
Sample size:
$$
n = \left(\frac{z_{1-\alpha} \sqrt{P_0(1-P_0)} + z_{1-\beta} \sqrt{P_a(1-P_a)}}{(P_0 - P_a)}  \right)^2
$$
Decision rule:
$$
d^* = nP_0 - z_{1-\alpha} \sqrt{nP_0(1-P_0)\left( \frac{N-n}{N-1}\right)}  
$$
We concluded that the *brexithealth* calculator does not use the use normal approximation to hypergeometric distribution in its calculation. 

**Input validation check:**

*check 1:* input: sample size = none (left empty) and other inputs are their default values, outputs: decision rule = 0, actual $\alpha$ error = 1, and actual $\beta$ error = 0. Calculator provides outputs even when sample size is missing.

*check 2:* input: sample size = 0 and other inputs are their default values, outputs: warning "Sample size is too small".

*check 3:* input: maximum tolerable $\alpha$ error = none (left empty) and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. Although the value of maximum tolerable $\alpha$ error is missing (not specified), the calculator provides outputs that are identical to outputs if maximum tolerable $\alpha$ error = 0.1.

*check 4:* input: maximum tolerable $\alpha$ error = 0 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\alpha$ error = 0.1.

*check 5:* input: maximum tolerable $\alpha$ error = 0.05 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\alpha$ error = 0.1.

*check 6:* input: maximum tolerable $\alpha$ error = 0.2 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\alpha$ error = 0.1.

*check 7:* input: maximum tolerable $\beta$ error = none (left empty) and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. Although the value of maximum tolerable $\alpha$ error is missing (not specified), the calculator provides outputs that are identical to outputs if maximum tolerable $\beta$ error = 0.1.

*check 8:* input: maximum tolerable $\beta$ error = 0 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\beta$ error = 0.1.

*check 9:* input: maximum tolerable $\beta$ error = 0.05 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\beta$ error = 0.1.

*check 10:* input: maximum tolerable $\beta$ error = 0.2 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0110, and actual $\beta$ error = 0.0057. The calculator provides outputs that are identical to outputs if maximum tolerable $\beta$ error = 0.1.

*check 11:* input: sample size = 40 and other inputs are their default values, outputs: decision rule = 26, actual $\alpha$ error = 0.0160, and actual $\beta$ error = 0.0161. Given sample size = 40 and decision rule = 26, the calculator generates correct actual $\alpha$ error and actual $\beta$ error.

*check 12:* input: upper threshold = 0.7 and other inputs are their default values, outputs: decision rule = 30, actual $\alpha$ error = 0.0757, and actual $\beta$ error = 0.0517. Shown below, the calculator generates correct outputs.

```{r}
find_dstar(N=N, p=0.7, n=50, alpha=0.1)
```
```{r}
# actual type 1 error
phyper(q=30, m = N*0.7, n = (N-N*0.7), k = 50)
```
```{r}
# actual type 2 error
1 - phyper(q=30, m = N*0.5, n = (N-N*0.5), k = 50)
```


*check 13:* input: upper threshold = 0.5 and other inputs are their default values, outputs: warning "Upper threshold must be above lower threshold".

*check 14:* input: upper threshold = none (left missing) and other inputs are their default values, outputs: decision rule = 0, actual $\alpha$ error = NaN, and actual $\beta$ error = 0. Calculator provides some outputs even when upper threshold is missing.

*check 15:* input: lower threshold = 0.4 and other inputs are their default values, outputs: decision rule = 30, actual $\alpha$ error = 0.0005, and actual $\beta$ error = 0.0009. Shown below, the calculator generates correct outputs.

```{r}
# actual type 1 error
phyper(q=30, m = N*0.8, n = (N-N*0.8), k = 50)
```

```{r}
# actual type 2 error
1 - phyper(q=30, m = N*0.4, n = (N-N*0.4), k = 50)
```


*check 16:* input: lower threshold = 0 and other inputs are their default values, outputs: decision rule = 0, actual $\alpha$ error = 0.000, and actual $\beta$ error = 0.000. Shown below, given decision rule = 0, the calculator generates correct outputs.

```{r}
# actual type 1 error
phyper(q=0, m = N*0.8, n = (N-N*0.8), k = 50)
```

```{r}
# actual type 2 error
1 - phyper(q=0, m = N*0, n = (N-N*0), k = 50)
```

*check 17:* input: population size = 1 and other inputs are their default values, outputs: warning "sample size must be smaller than population size"

*check 18:* input: population size = 1,000,000 and other inputs are their default values, outputs: decision rule = 33, actual $\alpha$ error = 0.0144, and actual $\beta$ error = 0.0077. Shown below, given decision rule = 33, the calculator generates correct outputs.

```{r}
# actual type 1 error
phyper(q=33, m = 1000000*p_0, n = (1000000-1000000*p_0), k = 50)
```

```{r}
# actual type 2 error
1 - phyper(q=33, m =1000000*p_a, n = (1000000-1000000*p_a), k = 50)
```







