---
title: "Review of LQAS Sampling Plan Calculator"
subtitle: "SQUEAC Calculator"
author: "Team 2"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Source: `SQUEAC.Setup.exe`

## 1. Summary

The *squeac* calculator calculates combinations of sample size and decision rule (decision threshold). Its input parameters are: upper standard (upper threshold proportion), lower standard (lower threshold proportion), lower error level (type 1 error), and upper error level (type 2 error). Given these input parameters, the calculator generates a list of pairs of sample size and its decision rule, and their associated actual lower error level (actual type 1 error) and actual upper error level (actual type 2 error).

Study type(s) this calculator would be suitable for:

* Point in time surveys to estimate a proportion

* Humanitarian surveys

* Comparative surveys


## 2. Review Checklist

1. What are the sampling assumptions  made for the calculator? 
* Simple random sampling without replacement
* Population size is very relative to the sample size such that computing sample sizes and their errors (type 1 and type 2) can be accomplished using normal distribution and binomial distribution, respectively. 
* One-sided test


2. What measure of accuracy is used for sample size calculation? 
* No mention of confidence interval
* No mention of coefficient of variation
* Other: upper error level and lower error level.


3. If it includes continuous variables how do they determine variance of estimates since a distribution needs to be known or assumptions made about the distribution?
* Not applicable because the calculator does not include continuous variables. 


4. Is the calculator calculation correct?

 Section 4 provides details of how we checked the calculator calculation.


5. If clustering of sample units, is a design effect suggested? What about for double sampling for stratification?
* The sampling unit is not cluster.
* No mention of design effect.
* The calculator only performs sampling plan calculation for single sampling.


6. Is probability proportional to size, stratification, double sampling?
* LQAS is identical to stratified sampling in which the samples are too small to provide acceptably narrow confidence intervals for estimates (Lemeshow and Stroh, 1989).


7. Accuracy of continuous variable estimate as well as that of proportion?
* The calculator only deals with proportion (in term of percentages).


8. How does calculator deal with effect of non-response bias on accuracy?
* No mention of non-response.


## 3. Limitations


## 4. Checking the calculator calculation

**Summary:**

### Verify calculation: input parameters set 1

To check the calculation of *squeac* calculator, we use input parameters below:

* lower standard: 50%

* upper standard: 80%

* upper error level: 10%

* lower error level: 10%

These input parameters are similar to those in Lemeshow and Taber (1991)

```{r}
# set input parameters
alpha <- 0.05 # upper error level
beta <- 0.2 # lower error level
p_0 <- 0.2    # upper standard
p_a <- 0.1    # lower standard

```

Suppose the calculator uses normal approximation to the hypergeometric distribution. Following Lemeshow and Taber (1991), the formula for sample size is: 

Sample size:
$$
n = \left(\frac{z_{1-\alpha} \sqrt{P_0(1-P_0)} + z_{1-\beta} \sqrt{P_a(1-P_a)}}{(P_0 - P_a)}  \right)^2
$$

```{r}
z_0 <- qnorm(alpha, lower.tail = FALSE)
paste0("z_0: ", round(z_0,3))
z_a <- qnorm(beta, lower.tail = FALSE)
paste0("z_a: ", round(z_a,3))
```

```{r}
# calculate n
numer <- z_0*sqrt(p_0*(1-p_0)) + z_a*sqrt(p_a*(1-p_a))
denom <- p_0 - p_a

n_calc <- ceiling((numer / denom)^2)
d_star <- n_calc*p_0 - (z_0 * sqrt(n_calc*p_0*(1-p_0)))

paste0("n: ", n_calc)
paste0("d_star: ", ceiling(d_star))
```

```{r}
# P(d<= 11) given n = 83 and d_star = 11 (results and input parameters as Lemeshow and Taber, 1991)
pbinom(11, size = 83, prob = 0.2) # also type 1 error
```
If the proportion of infected (or uncovered) people in the population is 0.2, the probability of observing 11 people infected in a sample of size 83 is 0.076 (about 7-8 percent), quite small, but larger than the specified upper error level of 5 percent. This is also the probability of concluding that the population has enough coverage (accept the lot) when in reality it is not. 

**check upper and lower given n = 81 and d = 10**

```{r}
# calculate type 1 error using calculator output: n = 81, d = 10.
# other input parameter: p_0 (upper standard) = 0.2
type_1_error <- pbinom(q=10, size = 81, prob = 0.2)
type_1_error # actual upper error level
```
The *squeac* calculator generates upper error = 5%. Thus, *squeac* output for upper error matches with our calculation of type 1 error = `r round(type_1_error, 3)`.

```{r}
# check type 2 error / "lower error" output, using calculator output: n = 81, d 10
# other input parameter: p_a (lower standard) = 0.1
1-pbinom(q=10, size = 81, prob = 0.1) # actual lower error level
```
The resulting calculation is similar to the output of *squeac* calculator of 18%.

```{r}
# recalculate type 1 error using normal distribution
pnorm(q = 10, mean = 81*p_0, sd = sqrt(81*p_0*(1-p_0))) # type 1 error
```

```{r}
# recalculate type 2 error using normal distribution
1 - pnorm(q = 10, mean = 81*p_a, sd = sqrt(81*p_a*(1-p_a)))

```

**check upper and lower given n = 82 and d = 10**
```{r}
# calculate type 1 error using calculator output: n = 82, d = 10.
# other input parameter: p_0 (upper standard) = 0.2
type_1_error <- pbinom(q=10, size = 82, prob = 0.2)
type_1_error # actual upper error level
```

calculator output, upper = 5%. very close. 

```{r}
#recalculate using normal distribution
pnorm(q = 10, mean = 82*p_0, sd = sqrt(82*p_0*(1-p_0))) # type 1 error
```
recalculation off.

```{r}
# check type 2 error / "lower error" output, using calculator output: n = 82, d 10
# other input parameter: p_a (lower standard) = 0.1
1-pbinom(q=10, size = 82, prob = 0.1) # actual lower error level
```

calculator output, lower = 19%. correct.

### Verify calculation: input parameters set 2

Check with different input parameters

```{r}
# set different input parameters
alpha <- 0.01 # upper error level
beta <- 0.2 # lower error level
p_0 <- 0.8    # upper standard
p_a <- 0.5    # lower standard
```

```{r}
z_0 <- qnorm(alpha, lower.tail = FALSE)
paste0("z_0: ", round(z_0,3))
z_a <- qnorm(beta, lower.tail = FALSE)
paste0("z_a: ", round(z_a,3))
```

```{r}
# calculate n
numer <- z_0*sqrt(p_0*(1-p_0)) + z_a*sqrt(p_a*(1-p_a))
denom <- p_0 - p_a

n_calc <- ceiling((numer / denom)^2)
d_star <- n_calc*p_0 - (z_0 * sqrt(n_calc*p_0*(1-p_0)))

paste0("n: ", n_calc)
paste0("d_star: ", ceiling(d_star))
```

calculator outputs start from n = 21 and d = 12. very close.

```{r}
# calculate type 1 error: n = 21, d = 13.
# other input parameter: p_0 (upper standard) = p_0 = 0.8
type_1_error <- pbinom(q=13, size = 21, prob = p_0)
type_1_error # actual upper error level
```

Higher than the specified $\alpha$, upper error level = 0.01

**check upper and lower given n = 21 and d = 12**

```{r}
# calculate type 1 error: n = 21, d = 13.
# other input parameter: p_0 (upper standard) = p_0 = 0.8
type_1_error <- pbinom(q=12, size = 21, prob = p_0)
type_1_error # actual upper error level
```

calculator output: 1%. correct.

```{r}
# check type 2 error / "lower error" output, using calculator output: n = 81, d 10
# other input parameter: p_a (lower standard) = 0.5
1-pbinom(q=12, size = 21, prob = p_a) # actual lower error level
```

calculator ouput: 19%. correct.
